What is an Ancestral Recombination Graph?

### 2022-03-24 paper.tex

Another limitation of the Griffiths ARG encoding illustated in
Fig.~\ref{fig-arg-data-structure}A is the inability to represent
multiple recombinations along a chromosome, gene conversion,
or transmission of multiple chromosomes.
[Sentence saying ``gene conversion is important, actually, cite, cite,
so we should be hoping to infer it some day''. Presumably people
already are in bacterial contexts?]
To allow us to encode data from these important processes we
must either devise workarounds to the event based encoding
(i.e., model gene conversion as two recombination events) or
generalise the model to more appropriately reflect the realities
of modern datasets.

Fortunately, it is straightforward to generalise the classical Griffiths
EARG encoding to both simplify implementations and to encompass any
possible pattern of genetic inheritance.
The only change that we require is to associate
the details of inheritance intervals with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}B
where this alternative edge annotation formulation is
shown. By associating the interval $(0, x)$ with edge $(3,4)$
and the interval $(x, L)$ with $(3, 5)$ we make the encoding
unambigous: we can list the edges in any order, and it
will describe the same set of local genealogical trees.
This approach is also more flexible, because any pattern
of inheritance can be described by these intervals. For
example, if a gene conversion event occured where a segment
between between positions $x$ and $y$ was inherited by
node $3$ from $5$, we would state this by associating
the intervals $\{(0, x], (y, L]\}$ with $(3,4)$ and
$\{(x, y]\}$ with $(3,5)$.
This change is straightforward, involves no loss of information
and (as we see in later sections)
greatly increases both the expressivity and computational
power of ancestral recombination graph data structures.

>>>>

Another limitation of the Griffiths ARG encoding illustrated in
Fig.~\ref{fig-arg-data-structure}A is the inability to represent
multiple recombinations along a chromosome, gene conversion,
or transmission of multiple chromosomes.
[Sentence saying ``gene conversion is important, actually, cite, cite,
so we should be hoping to infer it some day''. Presumably people
already are in bacterial contexts?]
To allow us to encode data from these important processes we
must either devise workarounds to the event based encoding
(i.e., model gene conversion as two recombination events) or
generalise the model to more appropriately reflect the realities
of modern datasets.

Fortunately, it is straightforward to generalise the classical Griffiths
EARG encoding to both simplify implementations and to encompass any
possible pattern of genetic inheritance.
The only change that we require is to associate
the details of inheritance intervals with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}B
where this alternative edge annotation formulation is
shown. By associating the interval $(0, x)$ with edge $(3,4)$
and the interval $(x, L)$ with $(3, 5)$ we make the encoding
unambiguous: we can list the edges in any order, and it
will describe the same set of local genealogical trees.
This approach is also more flexible, because any pattern
of inheritance can be described by these intervals. For
example, if a gene conversion event occurred where a segment
between between positions $x$ and $y$ was inherited by
node $3$ from $5$, we would state this by associating
the intervals $\{(0, x], (y, L]\}$ with $(3,4)$ and
$\{(x, y]\}$ with $(3,5)$.
This change is straightforward, involves no loss of information
and (as we see in later sections)
greatly increases both the expressivity and computational
power of ancestral recombination graph data structures.
